Why do languages tolerate heterography? An experimental investigation into the emergence of informative orthography

### 2024-01-19 manuscript/manuscript.tex
Respond to Reviewer 1 comments

One potential case of functional divergence is heterographic homophones---words that sound alike but which are written differently (e.g., \textit{meat} and \textit{meet}). Heterographic spellings such as these may serve a valuable function in reading. For example, faced with a spoken sentence beginning /ðɛr/, a listener will have high uncertainty about what word---or even what sentence structure---is likely to come next: a noun, as in /ðɛr kat/, a form of the verb \textit{to~be}, as in /ðɛr ɪz/, or the progressive form of a verb, as in /ðɛr gəʊɪŋ/. In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate these cases, giving the reader a headstart on processing the upcoming syntactic structure and semantic content. Heterography permits the written form of the language to be more informative than the spoken form. Although heterography is common in the more orthographically opaque languages, such as English, Hebrew, and Thai, even the most transparent of languages employ at~least some techniques to disambiguate homophones in writing. Italian, for example, uses the grave accent to distinguish common homophonous words: <la> (\textit{the}) vs.\@ <là> (\textit{there}).

@@

However, heterography comes with two main costs. Firstly, by definition, heterography implies that a single sound can be spelled multiple ways. The heterographic spellings <meat> and <meet>, for example, imply that the /i/ sound can be spelled <ea> or <ee>. Readers are therefore required to learn alternate spellings for a single sound, making alphabetic decoding more difficult \parencite{Seymour:2003, Spencer:2003, Taylor:2011, Zhao:2018}. Secondly, the mapping between heterographic forms and meaning must also be learned. From the point of view of the modern language user, there is no intrinsic reason why \textit{meat} is spelled <ea> and \textit{meet} is spelled <ee>; it could just as easily be the other way around. Nevertheless, these arbitrary spelling distinctions must be learned if they are to be useful, and they presumably place an additional burden on reading and---perhaps even more-so---on writing \parencite{Shankweiler:1992}.

@@

Participants were taught words for nine alien objects---three shapes (pentagon, star, torus) in three colors (pink, yellow, blue), as depicted in Fig.~\ref{stimuli}. The alien words had a spoken and written form composed of a stem and suffix. The stems, which always express the shape dimension, were /buvɪ/ <buvi> (the pentagon), /zɛtɪ/ <zeti> (the star), and /wɒpɪ/ <wopi> (the torus). These stems were fixed and unchanging throughout both experiments reported in this paper and were designed to be easy to learn by being graphically and phonetically iconic of the shapes they represent. Throughout Experiment~1, the spoken form of the suffix was always pronounced /kəʊ/, but its spelling was free to change over time. Thus, the spoken form of the language consists of just three unique words---/buvɪkəʊ/, /zɛtɪkəʊ/, and /wɒpɪkəʊ/---that mark only a shape distinction; however, the \textit{spelling} of the suffix could potentially take on different forms to mark color.

@@

Participants were arranged into transmission chains such that the spellings produced by one participant would subsequently be taught to the next participant in the chain (see Fig.~\ref{transmission}A). The first participant in a chain was taught the initial, randomly generated seed system, and this system was then free to evolve as it was transmitted to subsequent generations. Importantly, this process was subject to a bottleneck on transmission: Not all nine spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only six of the nine spellings produced at generation $i-1$ (at least one of each shape and at least one of each color). Nevertheless, participants were asked to produce a spelling for all nine objects, meaning that generalization was required for three unseen items. Transmission continued for nine generations in each of ten independent chains. In the Transmission~+ Communication condition (see Fig.~\ref{transmission}B), each generation consisted of a pair of participants, but the productions of only one of the two (the primary participant; determined by whichever participant started the experiment first) were iterated to the next generation. The productions of the secondary participant were not iterated any further and were thus a cultural deadend; the role of the secondary participant was to act as a genuine communicative partner for the primary participant, inducing pressure for the language to become informative.

@@

All participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials. Participants were also told explicitly that the stems were iconic of the objects' shapes, allowing participants to focus on learning the suffixes during the training phase. In passive exposure trials, the alien objects were presented alongside the written and spoken forms in quick succession for 2~s each. In mini-test trials, which were interleaved among the passive exposure trials, the participant was asked to type the appropriate written form for an object and was given feedback on any errors (deleted characters shown in red strikethrough text and additions shown in bold green text). The participant received a 2p bonus for spelling the word correctly but had to submit their response within 20~s. Each of the six object--word pairs in the training set (i.e., the seen items that passed through the bottleneck on transmission) was passively exposed 18 times and mini-tested six times, resulting in a total of 108 passive exposure trials and 36 mini-test trials (the maximum bonus in training was therefore 72p). To check that participants were listening to the spoken forms, they were asked auditorily to click on the alien object at three random points during training; participants who did not follow this instruction were excluded.

@@

After training, participants assigned to the Transmission-only condition completed a test phase, alternating between production and comprehension trials. In production trials, the participant was shown an object and heard its associated pronunciation. The participant's task was to type the appropriate spelling. The input box was limited to eight lowercase Latin characters, and participants had to spell the stem correctly to continue to the next trial. Since participants heard the word pronounced aloud, typing the stem correctly should have been trivial, but in cases where the stem was initially spelled incorrectly, a popup message explicitly reminded the participant of the correct spelling of the stem. This restriction was imposed to prevent the stems from diverging from their spoken forms over time; however, no such restriction was imposed on the spelling of the suffix. Since the overall word length was restricted to eight characters and since all stems were four letters long, participants could use, at a minimum, a zero suffix and, at a maximum, a four-letter suffix. In comprehension trials, the participant was shown a word and had to click on the matching object from an array of all nine objects arranged in random order (in cases where multiple objects were described by the same wordform, any of the objects was a valid choice). In both types of test trial, the participant was awarded a bonus of 2p for each correct answer, but no explicit feedback was provided. Each of the nine object--word pairs (i.e., including unseen items) was tested once in production and once in comprehension, resulting in 18 trials (the maximum bonus in test was therefore 36p).

@@

Participants assigned to the communicative condition completed a live, over-the-internet communication game with another participant, who had received training on the same orthographic system inherited from the previous generation (albeit subject to a separate application of the bottleneck on transmission; i.e., the seen items were selected independently for each participant). The game closely mirrored the overall structure of the test administered to participants in the non-communicative condition, with the production and comprehension trials becoming the two sides of a single communicative interaction. On a given trial, one participant (the director) would complete a production trial under the same input restrictions described above (i.e., produce a form for a target meaning), and the word they used was relayed to the other participant (the matcher), who would then complete a comprehension trial in response to that word (i.e., pick an object from the matcher array). The two participants then switched roles, resulting in the same overall trial structure as the Transmission-only condition (i.e., alternation between production and comprehension trials and the same bonusing scheme).

The framing and goal of the communication game was, however, quite different from the non-communicative test. In communication, the shared goal of the director and matcher is to have a successful communicative interaction, not necessarily to reproduce what they had learned in training. Both participants received the 2p bonus each time an interaction was successful; that is, the reward structure is not based on using the ``correct'' forms taught in training but based on accurately conveying meaning. The director is thus incentivized to produce a wordform that is unambiguous, and the matcher is incentivized to carefully interpret what the director has attempted to convey. The second important difference from the non-communicative test was that participants received rich feedback on the interaction: The director saw which object the matcher clicked on, and the matcher saw which object was the correct target. Overall, then, the communication game is identical to the non-communicative test in terms of the task to be performed (nine productions and nine comprehensions), but the goal is quite different. In the non-communicative test, the goal and reward structure is based on accurately reproducing the orthography learned during training, whereas in the communication game, the goal and reward structure is based on successfully communicating a target object.

@@

So, what was different about the present experiment compared the large body of prior experimental iterated learning studies? The primary difference was the presence of a spoken language that is decidedly \textit{not} informative about one of the dimensions. Indeed, the point of our experiment is to see whether orthography can resist phonology under sufficient pressure for informativeness. The presence of homophonous suffix forms acts as a cue to participants that the language itself does not mark color and that, therefore, the orthography should also not mark color. Faced with orthographic variation that could be conditioned on either dimension, learners appear to rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language. As a result, any variation in spelling comes to be associated with particular stems, resulting in the emergence of redundant suffix spellings that serve no real purpose. A rough analog of this redundancy can be found in English in the spelling of /-ʃən/ (<cian>, <cion>, <sion>, <ssion>, or <tion>), which is conditioned on the stem (e.g., \textit{magician} \textit{suspicion}, \textit{expulsion}, \textit{transmission}, and \textit{station}) following a complex set of rules \parencite[pp.~420--421]{Carney:1994}.

@@

Although the emergent orthographies tended to be slightly more informative under communicative pressure, suggesting a causal relationship between communicative need and informativeness in orthography, systematic differentiation was, on the whole, rare, unstable, and fleeting, be it through implicit generalization of the supplied variants or explicit innovation of new variants. This is \textit{not} because participants were unable to learn variant spellings; indeed, in many cases variant spellings were stable but ineffectually conditioned on shape rather than color. Nor was it because participants were unable to learn a system of color marking that is not expressed in the spoken language; we know from Experiment~2 that participants can learn such systems. Rather, it seems that, in Experiment~1, differentiation could not get of the ground because of two main factors. First, participants seem to have a strong bias against representing a meaning in writing when that meaning is not represented in speech. Put differently, participants seem to view spelling as subordinate to speech, so when asked to type in a word for a pink pentagon called /buvɪkəʊ/ to send to their partner, their natural inclination is to type a sequence of graphemes that transparently reflects that sequence of phonemes, not the meaning that those phonemes are attempting to express. Second, even when participants appreciate the need to differentiate the written forms to be successful, they often appear reluctant or unable to do so, perhaps because they view the written language as immutable or because the problem of aligning with a partner---even in a~synchronous setting---is too difficult to overcome. It is notable, for example, that in the communication games, many attempts to differentiate using English color letters were not reciprocated.

@@

Another limitation of this work is the extent to which lexical disambiguation really matters in real-world reading scenarios, since the syntactic and semantic context usually makes the meaning clear. If \textit{knight} and \textit{night} were spelled the same way, it is hard to imagine a context in which they might be confused. That being said, cultural transmission has been argued to have a strong amplifying effect on small cognitive biases \parencite{Thompson:2016}, so perhaps even a minor benefit in reading could have a large effect on orthography. An important issue in pursuit of this hypothesis will be to better understand the mechanism by which the biases of readers might place selective pressure on a writing system that is primarily shaped by the needs and preferences of writers.

It is also important to note that the functional explanation for heterography advanced here (i.e., ambiguity avoidance) is likely to be one of many. For example, \textcite{Stenroos:2016} take the view that English spelling has generally remained opaque with respect to phonology because its primary function was to be a record keeper across time and space. Written forms of languages resist change because, in terms of time, they need to be accessible across many decades or centuries, and, in terms of space, they need to be accessible across different jurisdictions and dialects.

>>>>

One potential case of functional divergence is heterographic homophones---words that sound alike but which are written differently (e.g., \textit{meat} and \textit{meet}). Heterographic spellings such as these may serve a valuable function in reading. For example, faced with a spoken sentence beginning /ðɛr/, a listener will have high uncertainty about what word---or even what sentence structure---is likely to come next: a noun, as in /ðɛr kat/, a form of the verb \textit{to~be}, as in /ðɛr ɪz/, or the progressive form of a verb, as in /ðɛr gəʊɪŋ/. In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate these cases, giving the reader a headstart on processing the upcoming syntactic structure and semantic content. Heterography permits the written form of the language to be more informative than the spoken form, \newmaterial{and this might be especially important given that written language often lacks extralinguistic cues to meaning, such as stress, context, and body language, as well as the fact that written language is often more sophisticated, increasing the pressure to avoid ambiguity}. Although heterography is common in the more orthographically opaque languages, such as English, Hebrew, and Thai, even the most transparent of languages employ at~least some techniques to disambiguate homophones in writing. Italian, for example, uses the grave accent to distinguish common homophonous words: <la> (\textit{the}) vs.\@ <là> (\textit{there}).

@@

However, heterography comes with two main costs. Firstly, by definition, heterography implies that a single sound can be spelled multiple ways. The heterographic spellings <meat> and <meet>, for example, imply that the /i/ sound can be spelled <ea> or <ee>. Readers are therefore required to learn alternate spellings for a single sound, making alphabetic decoding more difficult \parencite{Seymour:2003, Spencer:2003, Taylor:2011, Zhao:2018}. Secondly, the arbitrary mapping between heterographic forms and meaning must also be learned. From the point of view of the modern language user, there is no intrinsic reason why \textit{meat} is spelled <ea> and \textit{meet} is spelled <ee>; it could just as easily be the other way around. Nevertheless, these arbitrary spelling distinctions must be learned if they are to be useful, and they presumably place an additional burden on reading and---perhaps even more-so---on writing \parencite{Shankweiler:1992}. \newmaterial{Notably, opaque writing systems take longer to learn \parencite{} and present greater challenges for dyslexics \parencite{}.}

@@

Participants were taught words for nine alien objects---three shapes (pentagon, star, torus) in three colors (pink, yellow, blue), as depicted in Fig.~\ref{stimuli}. The alien words had a spoken and written form composed of a stem and suffix. The stems, which always express the shape dimension, were /buvɪ/ <buvi> (the pentagon), /zɛtɪ/ <zeti> (the star), and /wɒpɪ/ <wopi> (the torus). These stems were fixed and unchanging throughout both experiments reported in this paper and were designed to be easy to learn by being graphically and phonetically iconic of the shapes they represent \newmaterial{(e.g., the round torus shape is represented by ``round'' sounds/letters)}. Throughout Experiment~1, the spoken form of the suffix was always pronounced /kəʊ/, but its spelling was free to change over time. Thus, the spoken form of the language consists of just three unique words---/buvɪkəʊ/, /zɛtɪkəʊ/, and /wɒpɪkəʊ/---that mark only a shape distinction; however, the spelling of the suffix could potentially take on different forms to mark color.

@@

Participants were arranged into transmission chains such that the spellings produced by one participant would subsequently be taught to the next participant in the chain (see Fig.~\ref{transmission}A). The first participant in a chain was taught the initial, randomly generated seed system, and this system was then free to evolve as it was transmitted to subsequent generations. Importantly, this process was subject to a bottleneck on transmission: Not all nine spellings were transmitted from one generation to the next; rather, the participant at generation $i$ would observe only six of the nine spellings produced at generation $i-1$ (at least one of each shape and at least one of each color).\footnote{\newmaterial{The 2/3 bottleneck parameter was chosen based on common practices in the field and some piloting. In general, increasing this parameter will lead to a slower evolutionary process, so the value was chosen to allow for sufficient change to occur within nine generations.}} Nevertheless, participants were asked to produce a spelling for all nine objects, meaning that generalization was required for three unseen items. Transmission continued for nine generations in each of ten independent chains. In the Transmission~+ Communication condition (see Fig.~\ref{transmission}B), each generation consisted of a pair of participants, but the productions of only one of the two (the primary participant; determined by whichever participant started the experiment first) were iterated to the next generation. The productions of the secondary participant were not iterated any further and were thus a cultural deadend; the role of the secondary participant was to act as a genuine communicative partner for the primary participant, inducing pressure for the language to become informative.

@@

All participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials\newmaterial{, which lasted around 8~min}. Participants were also told explicitly that the stems \newmaterial{looked and sounded like} the objects' shapes, allowing participants to focus on learning the suffixes during the training phase. In passive exposure trials, the alien objects were presented alongside the written and spoken forms in quick succession for 2~s each. In mini-test trials, which were interleaved among the passive exposure trials, the participant was asked to type the appropriate written form for an object and was given feedback on any errors (deleted characters shown in red strikethrough text and additions shown in bold green text). The participant received a 2p bonus for spelling the word correctly but had to submit their response within 20~s. Each of the six object--word pairs in the training set (i.e., the seen items that passed through the bottleneck on transmission) was passively exposed 18 times and mini-tested six times, resulting in a total of 108 passive exposure trials and 36 mini-test trials (the maximum bonus in training was therefore 72p). To check that participants were listening to the spoken forms, they were asked auditorily to click on the alien object at three random points during training; participants who did not follow this instruction were excluded.

@@

After training, participants assigned to the Transmission-only condition completed a test phase, alternating between production and comprehension trials. In production trials, the participant was shown an object and heard its associated pronunciation. The participant's task was to type the appropriate spelling. The input box was limited to eight lowercase Latin characters, and participants had to spell the stem correctly to continue to the next trial. Since participants heard the word pronounced aloud, typing the stem correctly should have been trivial, but in cases where the stem was initially spelled incorrectly, a popup message explicitly reminded the participant of the correct spelling of the stem.\footnote{\newmaterial{Although artificial, these stem corrections were rarely encountered. 81\% of participants never encountered this message (they always spelled the stem correctly), 14\% encountered it once, and 5\% encountered it two or three times.}} This restriction was imposed to prevent the stems from diverging from their spoken forms over time; however, no such restriction was imposed on the spelling of the suffix. Since the overall word length was restricted to eight characters and since all stems were four letters long, participants could use, at a minimum, a zero suffix and, at a maximum, a four-letter suffix. In comprehension trials, the participant was shown a word and had to click on the matching object from an array of all nine objects arranged in random order (in cases where multiple objects were described by the same wordform, any of the objects was a valid choice). In both types of test trial, the participant was awarded a bonus of 2p for each correct answer, but no explicit feedback was provided \newmaterial{on the correctness of the signal or object selection}. Each of the nine object--word pairs (i.e., including unseen items) was tested once in production and once in comprehension, resulting in 18 trials (the maximum bonus in test was therefore 36p).

@@

Participants assigned to the communicative condition completed a live, over-the-internet communication game with another participant, who had received training on the same orthographic system inherited from the previous generation.\footnote{\newmaterial{Albeit subject to a separate application of the bottleneck on transmission (i.e., the seen items were selected independently for each participant). We felt this was more ecological than giving both participants identical input.}} \newmaterial{The communication game, which shares similarities with \textcite{Kirby:2015},} closely mirrored the overall structure of the test administered to participants in the non-communicative condition described above, with the production and comprehension trials becoming the two sides of a single communicative interaction. On a given trial, one participant (the director) would complete a production trial under the same input restrictions described above (i.e., produce a form for a target meaning), and the word they used was relayed to the other participant (the matcher), who would then complete a comprehension trial in response to that word (i.e., pick an object from the matcher array). The two participants then switched roles, resulting in the same overall trial structure as the Transmission-only condition (i.e., alternation between production and comprehension trials and the same bonusing scheme).

The framing and goal of the communication game was, however, quite different from the non-communicative test. In communication, the shared goal of the director and matcher is to have a successful communicative interaction, not necessarily to reproduce what they had learned in training. Both participants received the 2p bonus each time an interaction was successful; that is, the reward structure is not based on using the ``correct'' forms taught in training but based on accurately conveying meaning. The director is thus incentivized to produce a wordform that is unambiguous, and the matcher is incentivized to carefully interpret what the director has attempted to convey. The second important difference from the non-communicative test was that participants received rich feedback on the interaction: The director saw which object the matcher clicked on, and the matcher saw which object was the correct target. \newmaterial{As such, we view feedback as an intrinsic part of communicative interaction; thus, in the non-communicative test described above, no feedback was provided, as is the case in similar studies \parencite{Carr:2017, Motamedi:2019, Saldana:2019, Silvey:2019}.} Overall, the communication game is identical to the non-communicative test in terms of the task to be performed (nine productions and nine comprehensions), but the goal is quite different. In the non-communicative test, the goal and reward structure is based on accurately reproducing the orthography learned during training, whereas in the communication game, the goal and reward structure is based on successfully communicating a target object.

@@

So, what was different about the present experiment compared the large body of prior experimental iterated learning studies? The primary difference was the presence of a spoken language that is decidedly \textit{not} informative about one of the dimensions. Indeed, the point of our experiment is to see whether orthography can resist phonology under sufficient pressure for informativeness. The presence of homophonous suffix forms acts as a cue to participants that the language itself does not mark color and that, therefore, the orthography should also not mark color. Faced with orthographic variation that could be conditioned on either dimension, learners appear to rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language. As a result, any variation in spelling comes to be associated with particular stems, resulting in the emergence of redundant suffix spellings that serve no real purpose. \newmaterial{Similar outcomes have been noted before in the context of artificial language learning \parencite{Smith:2010},} and a rough analog can be found in English in the spelling of /-ʃən/ (<cian>, <cion>, <sion>, <ssion>, or <tion>), which is conditioned on the stem (e.g., \textit{magician} \textit{suspicion}, \textit{expulsion}, \textit{transmission}, and \textit{station}) following a complex set of rules \parencite[pp.~420--421]{Carney:1994}.

@@

Although the emergent orthographies tended to be slightly more informative under communicative pressure, suggesting a causal relationship between communicative need and informativeness in orthography, systematic differentiation was, on the whole, rare, unstable, and fleeting, be it through implicit generalization of the supplied variants or explicit innovation of new variants. This is \textit{not} because participants were unable to learn variant spellings; indeed, in many cases variant spellings were stable but ineffectually conditioned on shape rather than color. Nor was it because participants were unable to learn a system of color marking that is not expressed in the spoken language; we know from Experiment~2 that participants can learn such systems. Rather, it seems that, in Experiment~1, differentiation could not get of the ground because of two main factors. First, participants seem to have a strong bias against representing a meaning in writing when that meaning is not represented in speech. Put differently, participants seem to view spelling as subordinate to speech, so when asked to type in a word for a pink pentagon called /buvɪkəʊ/ to send to their partner, their natural inclination is to type a sequence of graphemes that transparently reflects that sequence of phonemes, not the meaning that those phonemes are attempting to express. \newmaterial{This has been described in other domains as ``functional fixedness'' \parencite{German:2000}; people might find it difficult to adduce a new function for orthography (expressing meaning) once they have become accustomed to another function (expressing sound).} Second, even when participants appreciate the need to differentiate the written forms to be successful, they often appear reluctant or unable to do so, perhaps because they view the spelling as immutable or because the problem of aligning with a partner---even in a~synchronous setting---is too difficult to overcome. It is notable, for example, that in the communication games, many attempts to differentiate using English color letters were not reciprocated.

@@

Another limitation of this work is the extent to which lexical disambiguation really matters in real-world reading scenarios, since the syntactic and semantic context usually makes the meaning clear. If \textit{knight} and \textit{night} were spelled the same way, it is hard to imagine a context in which they might be confused. That being said, cultural transmission has been argued to have a strong amplifying effect on small cognitive biases \parencite{Thompson:2016}, so perhaps even a minor benefit in reading could have a large effect on orthography. An important issue in pursuit of this hypothesis will be to better understand the mechanism by which the biases of readers might place selective pressure on a writing system that is primarily shaped by the needs and preferences of writers. \newmaterial{In many respects, writing systems are cultural fossils that do not readily adapt to external pressures.}

It is also important to note that the functional explanation for heterography advanced here (i.e., ambiguity avoidance) is likely to be one of many. For example, \textcite{Stenroos:2016} take the view that English spelling has generally remained opaque with respect to phonology because its primary function was to be a record keeper across time and space. Written forms of languages resist change because, in terms of time, they need to be accessible across many decades or centuries, and, in terms of space, they need to be accessible across different jurisdictions and dialects. \newmaterial{Another important thing to note is that the participant population we draw from (native English speakers) is already accustomed to heterography and opacity; the emergence of informative orthography might be less forthcoming among a population accustomed to transparency (e.g., participants whose native language is Finnish, Italian, or Serbian). We would nevertheless posit that informative orthography should still emerge in such populations, even if the process might be much slower (i.e., because the community first has to evolve the concept of directly encoding meaning). English, for example, was at one point much more transparent than it is now, but it has gradually switched to become more logographic. In this sense, our experiments jumpstart the process by adopting a participant population already used to heterography.}
